Peaks Over Threshold for Bursty Time Series

### 2018-04-20 article_springer/article_springer.Rmd
minor changes

  This article generalizes the Peaks Over Threshold (POT) model to the setting 
  where inter-event times are heavy-tailed and i.i.d.. For high thresholds and 
  infinite-mean waiting times, we show that the times between threshold 
  crossings are Mittag-Leffler distributed, and thus form a "fractional 
  Poisson Process" which generalizes the standard Poisson Process.
  We provide graphical means of estimating model parameters and assessing 
  model fit. Along the way, we apply our inference method to a real-world
  bursty time series, and 
  show how the Mittag-Leffler distribution has a memory which affects 
  the predictive distribution of the next extreme event.

@@

We review the marked renewal process in Section 2, and 
derive a scaling limit theorem for inter-exceedance times in Section 3.
We give a statistical procedure to estimate model parameters via stability plots 
in Section 5, but to set the stage we first discuss inference for the 
Mittag-Leffler distribution in Section 4. 
Diagnostic plots for model criticism are discussed in Section 6. 
In Section 7, we discuss the non-memorylessness property of the 
Mittag-Leffler distribution, and how it affects the predictive distribution 
for the next threshold crossing event. 
Section 8 concludes.
All statistical computations have been done in R, and all code, data, as 
well as the source code for this manuscript has been organized into an R 
software package, available at <https://github.com/strakaps/CTRM>.

@@

@@

In this setup, the $k$ magnitude $J_k$ occurs at time
$T_k = W_1 + \ldots + W_k$. 
Based on this simple model, we derive the following two random variables:

@@

Figure \ref{fig:thresholdedBursty} shows a simulated dataset in the top 
panel, where $W$ has a stable distribution with tail parameter 
$\beta =$ `r params$tail` (and skewness $1$ and location $0$), and 
where $J$ is from a standard Gumbel distribution. In the bottom panel, 
we plot a time series of solar flare intensities derived from a NASA
dataset [@HXRBS][^*]. Clearly, the simulated data exhibit long intervals 
_without any_ events, whereas in the real-world dataset events appear 
continuously. The threshold exceedances, however, appear to have the same 
statistical behaviour in both models. Observations below a threshold are 
commonly discarded in Extreme Value Theory; likewise, our MRP model
interprets these observations as noise and filters them out.

@@

In this section we state the key theorem below (a proof is in the appendix). 
For more details on regular variation and stable limit theorems, we 
recommend the book by @MeerschaertSikorskii.

@@

\noindent In general, for $\sigma > 0$ we write ${\rm ML}(\beta, \sigma)$ for the 
distribution of $\sigma W_\beta$. 
The Mittag-Leffler distribution with parameter $\beta \in (0,1]$ is 
a heavy-tailed positive distribution for $\beta < 1$, with infinite mean. 
However, as $\beta \uparrow 1$, ${\rm ML}(\beta, \sigma)$ converges 
weakly to the exponential distribution ${\rm Exp}(\sigma)$.
This means that although its moments are all infinite, the Mittag-Leffler
distribution may (if $\beta$ is close to 1) be indistinguishable from the 
exponential distribution, from the point of view of applied statistics. 
For a detailed reference on the Mittag-Leffler distribution, see e.g. 
@Haubold11.

@@

Since the Mittag-Leffler distribution is heavy-tailed, many researchers 
would intuitively give the highest importance to the tail behaviour of the 
distribution, and 
estimate the exponent of the tail function with established methods such 
as the Hill estimator. The QQ-estimator [@Kratz96] is closely related to 
the Hill estimator and fits a 
least squares line through the logarithms of the ordered statistics 
(y-axis) and the corresponding quantiles of the exponential distribution 
(x-axis), and returns the reciprocal slope as the estimate of the tail 
exponent.
For instance, the left panel in Figure \ref{fig:QQ-plots} shows the 
(static) QQ-estimator for the magnitudes of the solar flare data, returning 
a good fit to a Pareto distribution with tail parameter 
`r round(flares_exponent, 2)`.

However, as described nicely by @Resnick97, the less 
similar a heavy-tailed distribution is to a Pareto distribution, the less
useful a Hill estimator is. 
The top-right panel of \ref{fig:QQ-plots}, for instance, shows the 
QQ-estimator plot for $2000$ draws from the Mittag-Leffler distribution with 
tail parameter $0.8$. The stretched exponential shape means 
that the Mittag-Leffler distribution has more probability near $0$ than 
the Pareto distribution, severely biasing the estimator downwards 
(`r round(ML_exponent, 2)`). Hence a cutoff needs to be introduced, 
above which the tail exponent is fitted: The dynamic QQ-estimator 
plots the static QQ-estimates (y-axis) for different cutoffs (x-axis)
(see bottom-left panel). There is no apparent interval of stability, 
which could be taken to indicate a reasonable estimate for the true 
tail exponent $0.8$. 
Even worse, from a QQ-estimator point of view, there is no striking 
difference between the infinite mean ${\rm ML}(0.8,1)$ distribution and 
the light-tailed ${\rm Exp}(1)$ distribution (compare the right two panels). 
In particular, if only few (e.g. $<100$) data are available, 
then the static QQ-estimator plots look interchangeable for the two 
distributions.

>>>>

  This article generalizes the Peaks Over Threshold (POT) model to the setting 
  where inter-event times are heavy-tailed. For high thresholds and 
  infinite-mean waiting times, we show that the times between threshold 
  crossings are Mittag-Leffler distributed, and thus form a "fractional 
  Poisson Process" which generalizes the standard Poisson Process.
  We provide graphical means of estimating model parameters and assessing 
  model fit. Along the way, we apply our inference method to a real-world
  bursty time series, and 
  show how the Mittag-Leffler distribution has a memory which affects 
  the predictive distribution of the next extreme event.

@@

We review the marked renewal process in Section 2, and 
derive a scaling limit theorem for inter-exceedance times in Section 3.
We give a statistical procedure to estimate model parameters via stability plots 
in Section 5, but to set the stage we first discuss inference for the 
Mittag-Leffler distribution in Section 4. 
Diagnostic plots for model criticism are discussed in Section 6. 
In Section 7, we discuss the non-memorylessness property of the 
Mittag-Leffler distribution, and how it affects the predictive distribution 
for the next threshold crossing event. 
Section 8 concludes.
For all statistical computations we have used R [@R], and all code, data, as 
well as the source code for this manuscript has been organized into an R 
software package, available at <https://github.com/UNSW-MATH/CTRM>.

@@

In this setup, the $k$-th magnitude $J_k$ occurs at time
$T_k = W_1 + \ldots + W_k$. 
Based on this simple model, we derive the following two random variables:

@@

@@

Figure \ref{fig:thresholdedBursty} shows a simulated dataset in the top 
panel, where $W$ has a stable distribution with tail parameter 
$\beta =$ `r params$tail` (and skewness $1$ and location $0$), and 
where $J$ is from a standard Gumbel distribution. In the bottom panel, 
we plot a time series of solar flare intensities derived from a NASA
dataset [@HXRBS][^*]. Clearly, the simulated data exhibit long intervals 
_without any_ events, whereas in the real-world dataset events appear 
continuously. The threshold exceedances, however, appear to have similar
statistical behaviour in both models. Observations below a threshold are 
commonly discarded in Extreme Value Theory; likewise, our MRP model
interprets these observations as noise and filters them out.

@@

In this section we state and prove the key theorem, see below.
For an accessible introduction to regular variation and stable limit theorems, we 
recommend the book by @MeerschaertSikorskii.

@@

\noindent For a scale parameter $\sigma > 0$, we write 
${\rm ML}(\beta, \sigma)$ for the distribution of $\sigma W_\beta$. 
The Mittag-Leffler distribution with parameter $\beta \in (0,1]$ is 
a heavy-tailed positive distribution for $\beta < 1$, with infinite mean. 
However, as $\beta \uparrow 1$, ${\rm ML}(\beta, \sigma)$ converges 
weakly to the exponential distribution ${\rm Exp}(\sigma)$.
This means that although its moments are all infinite, the Mittag-Leffler
distribution may (if $\beta$ is close to 1) be indistinguishable from the 
exponential distribution, from the point of view of applied statistics. 
For a detailed reference on the Mittag-Leffler distribution, see e.g. 
@Haubold11.

@@

Since the Mittag-Leffler distribution is heavy-tailed, many researchers 
would intuitively give the highest importance to the tail behaviour of the 
distribution, and 
estimate the exponent of the tail function with established methods such 
as the Hill estimator. The QQ-estimator [@Kratz96] is closely related to 
the Hill estimator and fits a 
least squares line through the logarithms of the ordered statistics 
(y-axis) and the corresponding quantiles of the exponential distribution 
(x-axis). The reciprocal slope is returned as the estimate of the tail 
exponent.
For instance, the left panel in Figure \ref{fig:QQ-plots} shows the 
(static) QQ-estimator for the magnitudes of the solar flare data, returning 
a good fit to a Pareto distribution with tail parameter 
`r round(flares_exponent, 2)`.

However, as described nicely by @Resnick97, the less 
similar a heavy-tailed distribution is to a Pareto distribution, the less
useful a Hill estimator is. 
The top-right panel of Figure \ref{fig:QQ-plots}, for instance, shows the 
QQ-estimator plot for $2000$ draws from the Mittag-Leffler distribution with 
tail parameter $0.8$. The stretched exponential shape means 
that the Mittag-Leffler distribution has more probability near $0$ than 
the Pareto distribution, severely biasing the estimator downwards 
(`r round(ML_exponent, 2)`). Hence a cutoff needs to be introduced, 
above which the tail exponent is fitted: The dynamic QQ-estimator 
plots the static QQ-estimates (y-axis) for different cutoffs (x-axis)
(see bottom-left panel of Figure \ref{fig:QQ-est}). There is no apparent interval of stability, 
which could be taken to indicate a reasonable estimate for the true 
tail exponent $0.8$. 
Even worse, from a QQ-estimator point of view, there is no striking 
difference between the infinite mean ${\rm ML}(0.8,1)$ distribution and 
the light-tailed ${\rm Exp}(1)$ distribution (compare the right two panels). 
In particular, if only few (e.g. $<100$) data are available, 
then the static QQ-estimator plots look interchangeable for the two 
distributions.
